market-justice-preferences

### 2024-06-28 01-introduction.qmd

Given that the school environment has an important focus on performance, achievement and acknowledgment, meritocracy has been one of the principal concepts used for understanding and even for justifying performance differences among students. Meritocracy is a distributive system based on the belief that people should be rewarded and promoted based on their abilities, knowledge, and achievements [@young_rise_1958]. It is often seen as a way to create equal opportunities and fairness, as individuals can rise to positions of power and influence based on their own merit rather than their background or connections. However, some argue that meritocracy can actually lead to tolerating or even justifying social inequalities, as it can create a hierarchy where those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020;@mcnamee_meritocracy_2004]. In this regard, a great deal of academic research about meritocracy delves into the assessment of to what extent rewards and privileges in society are related to merit, emphasizing the so-called unfulfillable promise of meritocracy [@mijs_stratified_2016].

>>>>

Given that the school environment has an important focus on performance, achievement and acknowledgment, meritocracy has been one of the principal concepts used for understanding and even for justifying performance differences among students. Meritocracy is a distributive system based on the belief that people should be rewarded and promoted based on their abilities, knowledge, and achievements [@young_rise_1958]. It is often seen as a way to create equal opportunities and fairness, as individuals can rise to positions of power and influence based on their own merit rather than their background or connections. However, some argue that meritocracy can actually lead to tolerating or even justifying social inequalities, as it can create a hierarchy where those who already have resources and advantages are more likely to succeed [@sandel_tyranny_2020; @mcnamee_meritocracy_2004]. In this regard, a great deal of academic research about meritocracy delves into the assessment of to what extent rewards and privileges in society are related to merit, emphasizing the so-called unfulfillable promise of meritocracy [@mijs_stratified_2016].

### 2024-06-28 06-conclusion.qmd

The core of this paper was the association between meritocratic perceptions and preferences for market justice. In general, we find that those who perceive that effort and talent/intelligence are rewarded, are more willing to agree with that richer individuals can have better health, education, and pensions. This is not the first time that meritocratic perceptions (and also beliefs) have been related with the legitimation of social inequalities (Mijs, 2019; Darnon 2017), but the consideration of school level population as well as market justice preferences allows adding evidence regarding the role of meritocracy regarding inequality beliefs: legitimation of market criteria in social services seems to begin at school age, and what students think of meritocracy is linked to it. Although there are still several open questions, as for instance differences between meritocratic perceptions in school and in the society at large, this raises up the relevance of what is done (or not) in schools in order to challenge the meritocratic ideals.

>>>>

The core of this paper was the association between meritocratic perceptions and preferences for market justice. In general, we find that those who perceive that effort and talent/intelligence are rewarded, are more willing to agree with that richer individuals can have better health, education, and pensions. This is not the first time that meritocratic perceptions (and also beliefs) have been related with the legitimation of social inequalities [@mijs_paradox_2019; @darnon_where_2018], but the consideration of school level population as well as market justice preferences allows adding evidence regarding the role of meritocracy regarding inequality beliefs: legitimation of market criteria in social services seems to begin at school age, and what students think of meritocracy is linked to it. Although there are still several open questions, as for instance differences between meritocratic perceptions in school and in the society at large, this raises up the relevance of what is done (or not) in schools in order to challenge the meritocratic ideals.
